---
title: "Summarized_experiment_Jesica_quantile"
author: "Pérez Maiztegui, Alex"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```

# 14_LICA_Differential_Expression_Gene set enrichment analysis:
## Author: Alexperezm | Master's End of Degree Project - 2021-2022

Objective: Gene expression processing and classification. The survival rate has been split into low and high, based on the first and thrid terciles. The gene expression of those two groups will be studied in order to extract a curated list of genes directly related to the difference in survival rate. 

```{r}
setwd("C:/Users/Alex/OneDrive - alumni.unav.es/Master_Metodos_Computacionales_en_Ciencias/TFM/ClusteringJ_ANOVA/Dereg")
```

```{r}
library(SummarizedExperiment)
library(tidyverse)
library(readxl)
```

## 1st: Modify data:

```{r}
LICA <- read_excel("Jessica_subread_counts.xlsx")
survival <- read_excel("survival.xlsx", na = "NA")
tnorm <- data.frame(t(LICA))
colnames(tnorm) <- tnorm[1,]
tnorm <- slice(tnorm,2:162)
patient_id <- survival[,1]
```

```{r}
survival <- na.omit(survival)
surv <- survival[,2]
cuantiles<-quantile(surv, probs = c(.25,.50,.75,1),na.rm=TRUE)
```

```{r}
L_survival <- survival%>%
  filter(`Last survival (days)`<= cuantiles["25%"])
L_survival["Group"] <- "L"

H_survival <- survival%>%
  filter(`Last survival (days)`>= cuantiles["75%"])
H_survival["Group"] <- "H"

survival <- bind_rows(L_survival,H_survival)

tnorm["Patient"] <- rownames(tnorm)
LICA_survival <- inner_join(tnorm,survival,by="Patient")

traspuesto <- data.frame(t(LICA_survival))
write.csv(traspuesto,"LICA_58K_quantile_sumarized_experiment_type.csv")
```

```{r}
Jesica <- read_excel("LICA_58K_quantile_sumarized_experiment_type.xlsx")
G <- Jesica$GENE
Jesica <- Jesica%>%
  select(-GENE)
rownames(Jesica) <- G
Jesica <- Jesica[1:58288,]
nrows <- dim(Jesica)[1]
ncols <- length(Jesica)
rownames(Jesica) <- G[1:58288]

counts <- (Jesica)
counts <- as.matrix(counts)

rowRanges <- GRanges(rep(c("chr1"), 58288),
                     IRanges(floor(runif(58288, 1e5, 1e6)), width=100),
                     strand=sample(c("+", "-"), 58288, TRUE),
                     feature_id=list(Jesica[1:58288,1]),
                     row.names(Jesica[1:58288,1]))

grupos_sumarized_experiment_type <- read_excel("grupos_quantile_sumarized_experiment_type.xlsx")
grupos_sumarized_experiment_type["Patient"] <- str_replace(colnames(Jesica),"[:alpha:]","")
colData <- grupos_sumarized_experiment_type
```

```{r}
se <- SummarizedExperiment(assays=list(counts=counts), colData=colData,rowRanges = rowRanges)
saveRDS(se,"SummarizedExperiment_quantile_58k.rds")
```


## 2nd: Summarized experiment

```{r}
se <- readRDS("C:/Users/Alex/OneDrive - alumni.unav.es/Master_Metodos_Computacionales_en_Ciencias/TFM/ClusteringJ_ANOVA/Dereg/SummarizedExperiment_quantile_58k.rds")

colSums(assay(se))
head(assay(se), 3)
rowRanges(se)
str(metadata(rowRanges(se)))
colData(se)
class(se)
```

```{r}
library("DESeq2")

# Order of factors before relevel
se$Group
#se$Group <- relevel(se$Group, "L")


dds_raw <- DESeqDataSet(se, design = ~ Group)
colData(dds_raw)

head(assay(dds_raw))
rowRanges(dds_raw)
class(dds_raw)
```

#Visualization:
```{r}
head(counts(dds_raw))
rownames(dds_raw)
dds_raw
```

#pre-filter:
```{r}
keep <- rowSums(counts(dds_raw)) >= 50
dds <- dds_raw[keep,]


dds <- DESeq(dds)
res <- results(dds)
res
```

#Normalization:
```{r}
colSums(counts(dds_raw)) 
dds_raw <- estimateSizeFactors(dds_raw)
sizeFactors(dds_raw)

# counts()  allows to immediately retrieve the **normalized** read counts.
counts_normalized  <- counts(dds_raw, normalized = TRUE)
# Now take a look at the sum of the total depth after normalization 
colSums(counts(dds_raw, normalized=TRUE))

# Observe now the normalized counts
head(counts_normalized)
```


# Log^2^ normalization adding 1 pseudocount.
```{r}
counts_log_normalized <- log2(counts_normalized + 1)

par(mfrow=c(1,2)) # to plot  the  following  two  images side by side  each  other

# first , boxplots  of non -transformed  read  counts (one  per  sample)
boxplot(counts_normalized , notch = TRUE , las=2, cex.axis = 0.7,
        main = "untransformed  read  counts", ylab = "read  counts")

# box  plots  of log^2^ -transformed  read  counts
boxplot(counts_log_normalized , notch = TRUE , las=2, cex.axis = 0.7,
        main = "log2 -transformed  read  counts",
        ylab = "log2(read  counts)")
```

```{r}
library(vsn)
library(ggplot2)
library(hexbin)
```
# When the expected amount of variance is approximately the same across different mean values, the data is said to be homoskedastic. For RNA-seq counts, however, the expected variance grows with the mean.
```{r}
SdPlot <- meanSdPlot(counts_normalized, ranks = FALSE, plot = FALSE)  
SdPlot$gg + ggtitle("sequencing depth normalized") + ylab("standard deviation")
```

# The logarithm with a small pseudocount amplifies differences when the values are close to 0. The low count genes with low signal-to-noise ratio will overly contribute to sample-sample distances and PCA plots.
```{r}
SdPlot_log <- meanSdPlot(counts_log_normalized, ranks = FALSE, plot = FALSE)
SdPlot_log$gg + ggtitle("sequencing depth normalized log2(read counts)") + ylab("standard deviation")
```

# We can see this property of count with simulated data. Poison counts in a range of lambda 0.1 to 100
```{r}
lambda <- 10^seq(from = -1, to = 2, length = 1000)
cts <- matrix(rpois(1000*100, lambda), ncol = 100)
```

# Standard deviation of each row against the mean on raw counts
```{r}
sim_SdPlot <- meanSdPlot(cts, ranks = FALSE, plot = FALSE)
sim_SdPlot$gg + ggtitle("simulated data sequencing depth normalized)") + ylab("standard deviation")
```

# Standard deviation of each row against the mean on log2 pseudocount transformed data
```{r}
log.cts.one <- log2(cts + 1)
sim_SdPlot_log <- meanSdPlot(log.cts.one, ranks = FALSE, plot = FALSE)
sim_SdPlot_log$gg + ggtitle("simulated data sequencing depth normalized log2(read counts)") + ylab("standard deviation")
```

#Variance stabilizing transformation
```{r}
vsd <- vst(dds_raw, blind = FALSE)
head(assay(vsd), 3)
```

#rlog
```{r}
rld <- rlog(dds_raw, blind = FALSE)
head(assay(rld), 3)
```

```{r}
par(mfrow=c(2,2))
# Raw data normalized by sequencing depth
plot(counts_normalized [,1:2], cex=.1, main = "Normalized by sequencing depth")

# Log^2^ normalization adding 1 pseudocount.
plot(counts_log_normalized [,1:2], cex=.1, main = "Normalized log2(read counts)")

# rlog transformed
rlog_norm_counts <- assay(rld)
plot(rlog_norm_counts[,1:2], cex=.1, main = "rlog transformed", xlim=c(0,18), ylim=c(0,18))

# Variance stabilizing transformation
vsd_norm_counts <- assay(vsd)
plot(vsd_norm_counts[,1:2], cex=.1, main = "Variance stabilizing transformation", xlim=c(0,18), ylim=c(0,18))
```

#Clustering:
#Variance stabilizing transformation
```{r}
library(pheatmap)
library(RColorBrewer)
```

# Sample distances with VST
```{r}
sampleDists_vsd <- dist(t(assay(vsd)))
sampleDists_vsd
```

```{r}
# Transform sample distances to matrix
sampleDistMatrix_vsd <- as.matrix( sampleDists_vsd )
# Modify row names joining treatment and cell type
rownames(sampleDistMatrix_vsd) <- paste( vsd$dex, vsd$cell, sep = " - " )
# Remove col names
colnames(sampleDistMatrix_vsd) <- NULL
# Colors palette
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
```

# Draw heatmap
```{r}
heatmap <- pheatmap(sampleDistMatrix_vsd,
                    clustering_distance_rows = sampleDists_vsd,
                    clustering_distance_cols = sampleDists_vsd,
                    col = colors)
```

# Plot PCA on Type: Low / High Survival rate
```{r}
plotPCA(vsd, intgroup="Group")
```

#rlog:

# Sample distances with rld
```{r}
sampleDists_rld <- dist(t(assay(rld)))
sampleDists_rld
```

```{r}
# Transform sample distances to matrix
sampleDistMatrix_rld <- as.matrix( sampleDists_rld )
# Modify row names joining treatment and cell type
rownames(sampleDistMatrix_rld) <- paste( rld$dex, rld$cell, sep = " - " )
# Remove col names
colnames(sampleDistMatrix_rld) <- NULL
# Colors palette
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
```

# Draw heatmap
```{r}
heatmap <- pheatmap(sampleDistMatrix_rld,
                    clustering_distance_rows = sampleDists_rld,
                    clustering_distance_cols = sampleDists_rld,
                    col = colors)
```


# Plot PCA on Type
```{r}
plotPCA(rld, intgroup="Group")
```


#Differential gene expression

# Run DGE DeSeq2 pipeline
```{r}
dds_DGE <- DESeq(dds_raw)
```

# Mean-dispersion relationship 
```{r}
par(mfrow=c(1,1))
plotDispEsts(dds_DGE)
```

#Results table:
# Calling results() will build the base means across samples, log2 fold variation, standard errors, p value and p adjusted. 
```{r}
dds_DGE_results <- results(dds_DGE)
head(dds_DGE_results)
```

# Result object can be filter like a data frame. We have 4023 significant genes with a p adjusted < 0.05.  
```{r}
table(dds_DGE_results$padj < 0.05)
```

# We can also access to the metadata contained on the columns and check the pairwise contrast studied ()
```{r}
mcols(dds_DGE_results, use.names = TRUE)
```

# Summary of DGE
```{r}
summary(dds_DGE_results)
```

# Select significant genes with a p adjusted lower than 0.05
```{r}
resSig <- subset(dds_DGE_results, padj < 0.05)
```
# Order significant genes by p adjusted 
```{r}
resSig <- resSig[ order(resSig$padj), ]
```
# Order significant genes with the strongest down-regulation 
```{r}
head(resSig[ order(resSig$log2FoldChange), ])
```
# And with the strongest up-regulation
```{r}
head(resSig[ order(resSig$log2FoldChange, decreasing = TRUE), ])
```

```{r}
saveRDS(resSig, file="resSig_quantile_58k.rds")
saveRDS(rld, file="rld_quantile_58k.rds")

write.csv(resSig, "resultados_resSig.csv")
```


#ANNOTATE:

```{r}
#if (!require("BiocManager", quietly = TRUE))
#  install.packages("BiocManager")

#BiocManager::install("org.Hs.eg.db")

library("AnnotationDbi")
library("org.Hs.eg.db")
library(stringr)
```

```{r}
resSig_names <- rownames(resSig)
resSig_names <- str_sub(resSig_names,end = 15)

#resSig%>%mutate(names (resSig))
resSig$SYMBOL = mapIds(org.Hs.eg.db,
                       keys=resSig_names, #rownames(resSig)
                       column="SYMBOL",
                       keytype="ENSEMBL",
                       multiVals="first")

# Add gene name column
resSig$GENENAME = mapIds(org.Hs.eg.db,
                         keys=resSig_names, #rownames(resSig) 
                         column="GENENAME",
                         keytype="ENSEMBL",
                         multiVals="first")

# Add functional path column
resSig$PATH = mapIds(org.Hs.eg.db,
                     keys=resSig_names, #rownames(resSig)
                     column="PATH",
                     keytype="ENSEMBL",
                     multiVals="first")

# Add entrex ID column
resSig$ENTREZID = mapIds(org.Hs.eg.db,
                         keys=resSig_names, #rownames(resSig) 
                         column="ENTREZID",
                         keytype="ENSEMBL",
                         multiVals="first")
resSig$Type = mapIds(org.Hs.eg.db,
                     keys=resSig_names, #rownames(resSig)
                     column="GENETYPE",
                     keytype="ENSEMBL",
                     multiVals="first")

saveRDS(resSig,"resSig_anotado_cuantiles58k.rds")
```
#annotations_orgDb <- AnnotationDbi::select(org.Hs.eg.db, # database
#                                           keys = Jesica$,  # data to use for retrieval
#                                           columns = c("SYMBOL", "ENTREZID","GENENAME"), # information to retreive for given data
#                                           keytype = "ENSEMBL") # type of data given in 'keys' argument

# Add ENSEMBL column

```{r}
resSig$ENSEMBL <- rownames(resSig)

resSig_DF <- as.data.frame(resSig)
```

# Save significant DGE results to file
```{r}
write.table(resSig, file = "DESeq2_Sig_results_anotado_cuantiles58k.tab", sep = "\t", quote = FALSE , row.names = TRUE)
```

#Exploratory Plots:
```{r}
par(mfrow=c(1,1))
# Histogram of p-values frequencies  
hist(dds_DGE_results$pvalue , col = "blue",  xlab = "", , border = "white", ylab = "Frequency", breaks =0:40/40, main = "frequencies of p-values")
```

```{r}
par(mfrow=c(1,1))
# MA plot without shrinkage
#plotMA(dds_DGE_results, alpha = 0.05,  main = "Peritumor vs. Tumor", ylim = c(-5,5))

library("apeglm")
# Check possible contrast to create the MA plot
resultsNames(dds_DGE)

res <- lfcShrink(dds_DGE, coef="Group_L_vs_H", type="apeglm")
plotMA(res, alpha = 0.05, main = "L vs H log2 shrink",ylim = c(-5, 5))
```

#Volcano Plot:
```{r}
library("dplyr")

dds_DGE_results <- dds_DGE_results[order(dds_DGE_results$padj),]
results_order <- as.data.frame(dplyr::mutate(as.data.frame(dds_DGE_results), sig=ifelse(dds_DGE_results$padj<0.05, "FDR<0.05", "Not Sig")), row.names=rownames(dds_DGE_results))


p = ggplot2::ggplot(results_order, ggplot2::aes(log2FoldChange, -log10(padj))) +
  ggplot2::geom_point(ggplot2::aes(col = sig)) +
  ggplot2::scale_color_manual(values = c("red", "black")) +
  ggplot2::ggtitle("Significant genes Low vs High Survival")

p
```

#Heatmap:
# Select the 20 most variable genes
```{r}
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing = TRUE), 200)

# Transform to matrix and use vsd transformation
mat  <- assay(vsd)[ topVarGenes, ]
mat  <- mat - rowMeans(mat)
anno <- as.data.frame(colData(vsd)[, c("Group")])
rownames(anno) <- colnames(mat)
h <- pheatmap(mat, annotation_col = anno,show_rownames = FALSE,show_colnames = FALSE)
h
```

#GSEA:
```{r}
library(readxl)
#BiocManager::install("clusterProfiler", version = "3.8")
#BiocManager::install("pathview")
#BiocManager::install("enrichplot")
library(clusterProfiler)
library(enrichplot)
# we use ggplot2 to add x axis labels (ex: ridgeplot)
library(ggplot2)
library(readr)
library(tidyverse)
```

```{r}
setwd("C:/Users/Alex/OneDrive - alumni.unav.es/Master_Metodos_Computacionales_en_Ciencias/TFM/ClusteringJ_ANOVA/Dereg")
resultados_resSig <- read_csv("resultados_resSig.csv")
```

#Prepare input:
# we want the log2 fold change 
```{r}
original_gene_list <- resultados_resSig$log2FoldChange
```

# name the vector
```{r}
names(original_gene_list) <- resultados_resSig$...1
names(original_gene_list) <-  str_sub(names(original_gene_list),start = 1,end = 15)
```

# omit any NA values 
```{r}
gene_list<-na.omit(original_gene_list)
```

# sort the list in decreasing order (required for clusterProfiler)
```{r}
gene_list = sort(gene_list, decreasing = TRUE)
```

```{r}
organism = "org.Hs.eg.db"
#BiocManager::install(organism, character.only = TRUE)
library(organism, character.only = TRUE)

library(clusterProfiler)
```

```{r}
gse <- gseGO(geneList=gene_list, 
             ont ="ALL", 
             keyType = "ENSEMBL", 
             nPerm = 10000, 
             minGSSize = 3, 
             #maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = organism, 
             pAdjustMethod = "none")
resultados <- gse@result[,1:10]

library(writexl)
write_xlsx(gse@result)
```

#Dotplot:
```{r}
require(DOSE)
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)
```

#Enrichment map
```{r}
x2 <- pairwise_termsim(gse)
emapplot(x2,showCategory = 10,layout = "nicely")

emapplot(x2,
         showCategory = 10,
         color = "p.adjust",
         layout = "nicely",
         split = TRUE,
         pie = "equal",
         pie_scale = TRUE,
         cex_line = 1,
         min_edge = 0.2,
         cex_label_category = 1,
         node_label_size = TRUE,
)
```

#category network
# categorySize can be either 'pvalue' or 'geneNum'
```{r}
cnetplot(gse, categorySize="pvalue", foldChange=gene_list, showCategory = 6)
```

#ridgeplot:
```{r}
ridgeplot(gse,showCategory = 10) + labs(x = "enrichment distribution")
```

#GSEA plot:
# Use the `Gene Set` param for the index in the title, and as the value for geneSetId
```{r}
gseaplot(gse, by = "all", title = gse$Description[1], geneSetID = 1)
```

# KEGG Enrichment Analysis:

# Convert gene IDs for gseKEGG function
# We will lose some genes here because not all IDs will be converted
```{r}
ids<-bitr(names(original_gene_list), fromType = "ENSEMBL", toType = "ENTREZID", OrgDb=organism)
```

# remove duplicate IDS (here I use "ENSEMBL", but it should be whatever was selected as keyType)
```{r}
dedup_ids = ids[!duplicated(ids[c("ENSEMBL")]),]
```

# Create a new dataframe df2 which has only the genes which were successfully mapped using the bitr function above
```{r}
df2 = resultados_resSig[names(original_gene_list) %in% dedup_ids$ENSEMBL,]
```

# Create a new column in df2 with the corresponding ENTREZ IDs
```{r}
df2$trans = dedup_ids$ENTREZID
```

# Create a vector of the gene unuiverse
```{r}
kegg_gene_list <- df2$log2FoldChange
```

# Name vector with ENTREZ ids
```{r}
names(kegg_gene_list) <- dedup_ids$ENTREZID
```

# omit any NA values 
```{r}
kegg_gene_list<-na.omit(kegg_gene_list)
```

# sort the list in decreasing order (required for clusterProfiler)
```{r}
kegg_gene_list = sort(kegg_gene_list, decreasing = TRUE)
kegg_gene_list
```


kk2 <- gseKEGG(geneList     = kegg_gene_list,
               organism     = 'hsa',
               minGSSize    = 120,
               pvalueCutoff = 0.05,
               verbose      = FALSE)
head(kk2)
```